A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-09-10 main.tex
2020-09-10 meeting notes

Furthermore, the proportion of the environmental effect in the column space of $\boldsymbol{P}_{\bX}$ (i.e., the portion that will cause confounding) is given by 
\begin{equation}
    \label{eqn:ratio}
    \anna{||\bX \boldsymbol{\tau}||_2^2 / ||\bZ \bgamma||_2^2},
\end{equation}
\pb{should we have a symbol for this?}
which can range from 0 (no confounding) to 1 (entirely confounding). In the case where $\bZ \bgamma$ is entirely independent of population structure, $\bZ \bgamma \in \mathcal{N}(\boldsymbol{P}_{\bX})$ and $\bX \boldsymbol{\tau} = \boldsymbol{0}$ so that $\mathbb{E}(\bbetaHat^*) = \bbeta$. In this scenario, the SNP-phenotype associations will be unbiased, but the residual variance will be increased. Although we focus our attention on accurately estimating fixed SNP effects, we note that this result could lead to substantial bias in narrow-sense heritability estimates. \pb{Hmm...could this sentence be moved somewhere else?  I feel that narrow-sense heritability deserves a paragraph somewhere}

@@

\subsubsection{Subpopulation structure}
For the fully simulated data, two subpopulation structures were considered, which we refer to as ``fine'' and ``coarse''.  We define a coarse structure to be one characterized by a few large subpopulations intended to model large-scale environmental heterogeneity and population stratification, such as that described in our motivating air quaility example. Conversely, we define a fine subpopulation structure to be one characterized by many small subpopulations, representing a more subtle continuum of environmental heterogeneity \pb{do you know, are the subpopulations more similar in the fine case?}. Whether fine or coarse, the simulated data were generated such that subpopulations (also known as founding populations) are of equal size. We note that the distinction between fine and coarse structure is an entirely different concept than the level of admixture in the simulated data. For example, the discrete nonoverlapping subpopulations in the simulated Independent Subpopulations data result in an RRM which more directly corresponds to the discrete and fixed environmental effects, fine or coarse, than do the continuous and overlapping subpopulations in the simulated 1D Linear Admixture data \pb{?????}. The Semi-Empirical data is comprised of three racial subpopulations; one relatively large, one small, and one in the middle. Thus it represents a hybrid of our simulated coarse and fine structures, with little apparent admixture. Supplementary Figures \anna{?, ?, and ? MAKE THESE AND MAKE THEM PRETTY} illustrate the simulated kinship structures in the fine and coarse subpopulation settings, as well as that present in the empirical SNP data.

@@

@@

\subsubsection{Tuning parameter selection}
Selection of the tuning parameter $\lambda$ plays an important role in all penalized regression methods, with many possible approaches.  Here, for each model, we selected the $\lam$ value such that 50 SNPs were selected (recall that 50 is also the true number of causal SNPs). This method for selecting $\lambda$ was chosen in an attempt to compare the performance of all methods fairly, and as a strategy to evaluate methods in terms of variable selection, as it allows for direct comparison of the first 50 variables each method lets into the model. Alternative $\lambda$ selection methods were also considered. \pb{This seems way too strong}  Cross-validation has been found to be inadvisable for LMMs, and in the presence of non-inedpendent samples as is the case here. While information criterion methods such as BIC present another option for selecting $\lambda$, they may be unstable in high dimensions. In addition, because Lasso, PC-Lasso, and LMM-Lasso use different likelihoods, an information-criterion-based approach would not lend itself to comparable results across methods.

@@

>>>>

Furthermore, the proportion of the environmental effect in the column space of $\boldsymbol{P}_{\bX}$ (i.e., the portion that will cause confounding) is given by 
\begin{equation}
    \label{eqn:ratio}
    \anna{||\bX \boldsymbol{\tau}||_2^2 / ||\bZ \bgamma||_2^2},
\end{equation}
\pb{should we have a symbol for this?}
which can range from 0 (no confounding) to 1 (entirely confounding). In the case where $\bZ \bgamma$ is entirely independent of population structure, $\bZ \bgamma \in \mathcal{N}(\boldsymbol{P}_{\bX})$ and $\bX \boldsymbol{\tau} = \boldsymbol{0}$ so that $\mathbb{E}(\bbetaHat^*) = \bbeta$. In this scenario, the SNP-phenotype associations will be unbiased, but the residual variance will be increased. Although we focus our attention on accurately estimating fixed SNP effects, we note that this result could lead to substantial bias in narrow-sense heritability estimates. \pb{Hmm...could this sentence be moved somewhere else?  I feel that narrow-sense heritability deserves a paragraph somewhere} \anna{move to discussion do a paragraph on narrow-sense heritability... reference this eq - and give it a label}

@@

\subsubsection{Subpopulation structure}
For the fully simulated data, two subpopulation structures were considered, which we refer to as ``fine'' and ``coarse''.  We define a coarse structure to be one characterized by a few large subpopulations intended to model large-scale environmental heterogeneity and population stratification, such as that described in our motivating air quaility example. Conversely, we define a fine subpopulation structure to be one characterized by many small subpopulations, representing a continuum of environmental heterogeneity. Whether fine or coarse, the simulated data were generated such that subpopulations (also known as founding populations) are of equal size. We note that the distinction between fine and coarse structure is an entirely different concept than the level of admixture in the simulated data. \anna{size of the subpopulations that determines fine/course - cut into 2 sentences. similarly for the 1d amixture...} For example, the discrete nonoverlapping subpopulations in the simulated Independent Subpopulations data result in an RRM which more directly corresponds to the discrete and fixed environmental effects, fine or coarse, than do the continuous and overlapping subpopulations in the simulated 1D Linear Admixture data \pb{?????}. The Semi-Empirical data is comprised of three racial subpopulations; one relatively large, one small, and one in the middle. Thus it represents a hybrid of our simulated coarse and fine structures, with little apparent admixture. Supplementary Figures \anna{?, ?, and ? MAKE THESE AND MAKE THEM PRETTY} illustrate the simulated kinship structures in the fine and coarse subpopulation settings, as well as that present in the empirical SNP data.

@@

\anna{complex/noncomplex ... can be complex in different ways... discordant vs concordant ... dichotomous vs exponential}

@@

\subsubsection{Tuning parameter selection}
Selection of the tuning parameter $\lambda$ plays an important role in all penalized regression methods, with many possible approaches.  Here, for each model, we selected the $\lam$ value such that 50 SNPs were selected (recall that 50 is also the true number of causal SNPs). This method for selecting $\lambda$ was chosen in an attempt to compare the performance of all methods fairly, and as a strategy to evaluate methods in terms of variable selection, as it allows for direct comparison of the first 50 variables each method lets into the model. \anna{make this sentence more getting at the point that we had concerns about using these common lambda selection methods... Although CV and information criterion-based methods are commonly used to select $\lambda$ values, in this context they have the potential to result in misleading comparisons across methods.} Alternative $\lambda$ selection methods were also considered. \anna{I know i have a citation for this somewhere FIND IT} Cross-validation may be inadvisable for LMMs, and in the presence of non-inedpendent samples as is the case here. While information criterion methods such as BIC present another option for selecting $\lambda$, they may be unstable in high dimensions. In addition, because Lasso, PC-Lasso, and LMM-Lasso use different likelihoods, an information-criterion-based approach would not lend itself to comparable results across methods.

@@

\anna{check this value for sampsizes of 588 for admixture and independent...maybe even diff subpop sizes? at least to pin down the issue}
